---
title: "Git and GitHub"
# linktitle: "1. Introduction to RStudio and Images"
date: "2025-08-27"
class_date: "2025-08-27"
draft: false
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 2
type: docs
weight: 2
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Git

<a href="../../slides/01c_Git/01c_Git.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01c_Git/01c_Git.html#1')
```

## GitHub

<a href="../../slides/01c_Git/01d_GitHub.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01c_Git/01d_GitHub.html#1')
```

## SSH

<a href="../../slides/01c_Git/01e_SSH.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01c_Git/01e_SSH.html#1')
```

## References

- One-pager simple `git` guide. [https://rogerdudler.github.io/git-guide/](https://rogerdudler.github.io/git-guide/) 

- One-pager of `git` commands. [https://github.com/kbroman/Tools4RR/blob/master/04_Git/GitCommands/git_notes.md](https://github.com/kbroman/Tools4RR/blob/master/04_Git/GitCommands/git_notes.md) 

- Learn `git` interactively in 15 min. [https://try.github.io/levels/1/challenges/1](https://try.github.io/levels/1/challenges/1) 

- Interactive git branching tutorial. [http://learngitbranching.js.org/](http://learngitbranching.js.org/) 

- "Git and GitHub guide" by Karl Broman [http://kbroman.org/github_tutorial/](http://kbroman.org/github_tutorial/) 

- Software Carpentry course on `git`. [https://swcarpentry.github.io/git-novice/](https://swcarpentry.github.io/git-novice/) 

- Blischak, John D., Emily R. Davenport, and Greg Wilson. “A Quick Introduction to Version Control with Git and GitHub.” Edited by Francis Ouellette. PLOS Computational Biology 12, no. 1 (January 19, 2016): e1004668. [https://doi.org/10.1371/journal.pcbi.1004668](https://doi.org/10.1371/journal.pcbi.1004668) - An excellent explanation of Git and GitHub. Definitions (Box 1), tutorial

- Bryan, Jennifer. “Excuse Me, Do You Have a Moment to Talk about Version Control?” [https://doi.org/10.7287/peerj.preprints.3159v2](https://doi.org/10.7287/peerj.preprints.3159v2)

- Book(down) "Happy Git and GitHub for the useR" by Jenny Bryan. [http://happygitwithr.com/](http://happygitwithr.com/) 

- How to create pull requests. [https://akrabat.com/the-beginners-guide-to-contributing-to-a-github-project/](https://akrabat.com/the-beginners-guide-to-contributing-to-a-github-project/) 

- Quick Git and GitHub videos. [http://www.dataschool.io/git-and-github-videos-for-beginners/](http://www.dataschool.io/git-and-github-videos-for-beginners/) 

