---
title: "Differential Expression"
# linktitle: "1. Introduction to RStudio and Images"
date: "2025-10-06"
class_date: "2025-10-06"
draft: false
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 6
type: docs
weight: 6
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Differential expression

<a href="../../slides/07a_DiffExpr/07a_Intro.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/07a_Intro.html#1')
```

<a href="../../slides/07a_DiffExpr/02_RNA-seq_expdesign.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/07b_design_matrices.html#1')
```

<a href="../../slides/07a_DiffExpr/07b_design_matrices.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/07c_multipletesting.html#1')
```

<a href="../../slides/07a_DiffExpr/07c_multipletesting.html#1" target="_blank">Lecture slides in HTML format</a>


```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/07d_Overview_stats.html#1')
```

<a href="../../slides/07a_DiffExpr/07d_Overview_stats.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/07e_limma.html#1')
```

<a href="../../slides/07a_DiffExpr/07e_limma.html#1" target="_blank">Lecture slides in HTML format</a>


## Normalization

<a href="../../slides/07a_DiffExpr/05_RNA-seq_normalization.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/05_RNA-seq_normalization.html#1')
```

<a href="../../slides/07a_DiffExpr/07a_DiffExpr_batch.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/07a_DiffExpr/07a_DiffExpr_batch.html#1')
```

### Code examples

- Demo00_log2, <a href="../../slides/07a_DiffExpr/lab/Demo00_log2.Rmd" target="_blank">Rmd</a>, <a href="../../slides/07a_DiffExpr/lab/Demo00_log2.html" target="_blank">HTML</a>
- Demo01_loess, <a href="../../slides/07a_DiffExpr/lab/Demo01_loess.Rmd" target="_blank">Rmd</a>, <a href="../../slides/07a_DiffExpr/lab/Demo01_loess.html" target="_blank">HTML</a>
- Demo02_quantile_norm, <a href="../../slides/07a_DiffExpr/lab/Demo02_quantile_norm.Rmd" target="_blank">Rmd</a>, <a href="../../slides/07a_DiffExpr/lab/Demo02_quantile_norm.html" target="_blank">HTML</a>
- Demo03_MA_TMM, <a href="../../slides/07a_DiffExpr/lab/Demo03_MA_TMM.Rmd" target="_blank">Rmd</a>, <a href="../../slides/07a_DiffExpr/lab/Demo03_MA_TMM.html" target="_blank">HTML</a>
- Demo04_ComBat_SVA, <a href="../../slides/07a_DiffExpr/lab/Demo04_ComBat_SVA.Rmd" target="_blank">Rmd</a>, <a href="../../slides/07a_DiffExpr/lab/Demo04_ComBat_SVA.html" target="_blank">HTML</a> (<a href="../../slides/07a_DiffExpr/lab/Analysis_GEO.zip" target="_blank">Data preparation scripts</a>) 
-->
